---
title: 'Application Activity:  Descriptive Statistics'
format:
  html:
    self-contained: true
    embed-resources: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```



# Introduction

In this activity, you will practice everything we've covered up to this point including:

* Reviewing a dataset  
* Determining response and explanatory variables  
* Determining data types (quantitative, categorical)  
* Creating descriptive statistics:  
  + Quantitative data (one and two variables)
    + Summary statistics for a single variable (`favstats(data$y)`)
    + Histogram for one variable (`histogram(data$y)`)  
    + Summary statistics for multiple groups (`favstats(data$y ~ data$x)`)  
    + Side-by-side boxplots (`boxplot(data$y ~ data$x)`)  
    + Scatter plots, continuous $x$ (`plot(data$y ~ data$x)`)  
    + Correlation between 2 quantitative variables (correlation coefficient, $r$)  
  + Categorical data (one and two variables)  
    + Identifying levels of a categorical variable (`unique()`)  
    + Frequency and proportion tables (`table(data$variable)`, `prop.table(table(data$variable))`)  
    + Univariate, sorted bar charts (`barplot(sorted_table_name)`)  
    + Contingency tables (`table(row, column)`)  
    + Row/Column Percents (`prop.table(tbl_name, margin = )`)  
    + Bivariate bar charts (`barplot(table_name, beside=TRUE, col=c(2,3,4,...))`)  
    
  
We will be using data collected about students in 2 Portuguese schools including their final grade. The goal of the research was to see what variables impact final grades.  

## Load the Libraries and Data

```{r, echo=FALSE, include=FALSE}

## Don't worry about this line of code. It is included for purposes of rendering the document

library(rio)
library(tidyverse)
library(mosaic)
library(car)

student <- read_csv('https://raw.githubusercontent.com/byuistats/Math221D_Cannon/master/Data/student_data_kaggle.csv')

```

# Review the data

Take some time to familiarize yourself with the data.  Check the [website](https://www.kaggle.com/datasets/whenamancodes/student-performance) familiarize yourself with the study.  See what the columns are.  


## Response Variable:  Univariate Summary Statistics

__QUESTION__:  What is the primary response variable?  (What are we trying to predict?)  
__ANSWER__:  

__QUESTION__:  Create a table of summary statistics for the primary response variable:  

```{r}

favstats()

```

__QUESTION__:  What is the mean grade?  
__ANSWER__:  

__QUESTION__:  What is the median grade?  
__ANSWER__:  

__QUESTION__:  What is the Standard Deviation of grades?  
__ANSWER__:  

__QUESTION__:  Interpret in words what the standard deviation means?  
__ANSWER__:  

__QUESTION__:  What is the 75th percentile of grades?  
__ANSWER__:  

__QUESTION__:  Interpret in words what the 75th percentile means?  
__ANSWER__:  


Create a histogram of the primary response variable:  

```{r}



```

__QUESTION__:  What is the basic shape of the distribution?  
__ANSWER__:  

__QUESTION__:  What anomalies, if any, do you notice?  
__ANSWER__:  



## Explanatory Variables

List 5 potential explanatory variables and whether or not they are *quantitative* or *categorical*:  

  1.  

  2.  
 
  3.  

  4.  

  5.  


### Relationship Between 2 Variables

#### Categorical X

Create a table of summary statistics for final grade for each level of `sex`:  

```{r}

favstats(~)

```

__QUESTION__:  What is the mean grade for females?  
__ANSWER__:  


Create a side-by-side boxplot of final grades for each level of father's job (`Fjob`):

```{r}


```

__QUESTION__:  Which level of Fjob tends to have students with the highest final grades?  
__ANSWER__:  



Create a side-by-side boxplot of final grades for each level of mother's job (`Mjob`):

```{r}


```

__QUESTION__:  Which level of Mjob appears to have students with the highest final grades?  
__ANSWER__:  




#### Quantitative X

Create a scatter plot looking at the relationship between mid-term grade (G2) and final grade (G3):

```{r}


```

__QUESTION__:  Does the relationship appear roughly linear?  
__ANSWER__:  

__QUESTION__:  What anomalies, if any, do you notice?  
__ANSWER__:  

__QUESTION__:  Based on the scatter plot, guess the correlation coefficient:    
__ANSWER__:  

Calculate the correlation coefficient, r: 

```{r}

```





## Categorical Descriptive Statistics


__QUESTION__:  What are the levels of mother's profession?  

```{r}

unique()

```


Create a proportion table table for `sex`

```{r}


```

__QUESTION__:  What percent of respondents are female?  
__ANSWER__:  



Create a sorted bar chart of the counts of reasons students chose their school (`reason`):  

```{r}



```

__QUESTION__:  Which reason appears to be the most popular?  
__ANSWER__:  



Create and name a contingency table looking at the relationship between gender and whether or not they are in a romantic relationship.  

Then calculate the proportion of each gender in a romantic relationship (should sum to 1 for each gender):  

```{r}


```


__QUESTION__:  What percent of female respondents are in a romantic relationship?  
__ANSWER__:  


__QUESTION__:  What percent of male respondents are in a romantic relationship?  
__ANSWER__:  


Create a bar chart that groups bars based on `sex` and colors the bars by `romantic`.

```{r}

# Start by naming the table
tbl_relationship <- table()

# Then create the chart:  

```






