---
title: "Confidence Intervals and Hypothesis Tests Practice"
subtitle: "Proportions"
---


# Test Requirements

Recall that we can use the normal distribution to approximate the distribution of $\hat{p}$ under certain conditions.  

__Question__:  What to we need to check before we assume normality for a *hypothesis test*?  
__Answer__:  

__Question__:  What to we need to check before we assume normality for *confidence intervals*?  
__Answer__:  


# Teenage Smoking Habits

A study was conducted to determine the proportion of American teenagers between 13 and 17 who smoke. Previous surveys showed that 15% percent of all teenagers smoke. A Gallup survey interviewed a nationally representative sample of 785 teenagers aged 13 to 17. Seventy-one (71) teenagers in the survey acknowledged having smoked at least once in the past week. 

We want to explore if the study provide adequate evidence to conclude that the percentage of teenagers who smoke is now *different* than 15%.

__Question__:  Are the requirements for both a confidence interval and a hypothesis test met and why?  
__Answer__:  


Whether or not the test requirements are met, perform the appropriate hypothesis test and create a confidence interval for the true proportion of teenagers who smoke.

## Hypothesis Test

State the null and alternative hypotheses (replace the ??? with the correct information):

$$H_0: p=???$$  
$$H_a: p???0.15$$
Choose your confidence level:

$$\alpha = 0.$$  

```{r}

```

__Question__:  What is the value of the test statistic?  
__Answer__:  

__Question__:  What is the P-value?  
__Answer__:  

__Question__:  Based on your selected $\alpha$ and P-value, what is your conclusion?  
__Answer__:  


## Confidence Interval

Create a 99\% confidence interval for the true population proportion of teenagers who smoke?  

```{r}

```



## Proportion of Left-handed Artists

Suppose you would like to study creativity of left-handed people.  Your favorite generative AI told you that the estimated population of left-handed people in the United States is 11\%. You sample 122 Visual Arts majors and find that 20 are left handed.  Is this enough evidence to demonstrate that visual arts majors have a higher proportion of left handed people than the general population?  

__QUESTION__:  What is the estimated $\hat{p}$?  
__ANSWER__:  


__QUESTION__:  Are the requirements for both a hypothesis test met?  Why?  
__ANSWER__:  


```{r}

```


Perform a hypothesis test to determine if the proportion of left-handed Visual Art majors is higher than the general average of 11%.  

```{r}


```

__Question__:  What is the value of the test statistic?  
__Answer__:  

__Question__:  What is the P-value?  
__Answer__:  

__Question__:  Based on the P-value and a confidence level of $\alpha=.05$, what is your conclusion?  
__Answer__:  



## Confidence Interval

Create a 95\% confidence interval for the true population proportion of teenagers who smoke?  

```{r}

```


__QUESTION__:  Interpret the confidence interval in context of the research question:  
__ANSWER__:  



# Bird Watching

A wildlife researcher is studying a rare bird species in a southeast Idaho.  They identify a sample of 500 birds, and only 6 are found to be of this rare species. 


Estimate the proportion of the entire bird population that is this rare species using a 95% confidence interval.  

```{r}
prop.test(x = , n = , conf.level=)$conf.int
```


__QUESTION__:  Interpret the confidence interval in context of the research question:  
__ANSWER__:  


__QUESTION__:  Are the assumptions for the normality of $\hat p$ are met?  (Check $n\hat{p}$ and $n(1-\hat{p})$)  
__ANSWER__:  

__QUESTION__:  Perform a statistical hypothesis test to see if the population proportion, $p$, is less than 3%.  Use $\alpha=0.05$.  

```{r}

prop.test(x=, n=, p=, alternative=)

```


__QUESTION__:  What is the p-value for the above test?  
__ANSWER__:  


__QUESTION__:  State your conclusion using the technical requirements and in context of the research question.  
__ANSWER__:  

__QUESTION__:  Are the test requirements for a statistical hypothesis for a proportion met?  (Check $np$ and $n(1-p)$)  
__ANSWER__:  

