---
title: Confidence Intervals For Means - Practice
format:
  html:
    self-contained: true
    embed-resources: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```



# Introduction

In this exercise, you will create several confidence intervals for an unknown __population mean__, $\mu$, based on your sample mean, $\bar x$.  

Start by loading the libraries: 

```{r}

library(rio)
library(mosaic)
library(tidyverse)
library(car)

```



# Creativity

Create a 95% confidence interval for the __true population average__ creativity percentile (`Creativity_Percentile`) of Brother Cannon's students ($\mu_{\text{creativity}}$):  

```{r}

creativity <- read_csv('https://github.com/byuistats/Math221D_Course/raw/main/Data/creativity_personality_Clean.csv')

t.test(, conf.level = )$conf.int
```

__QUESTION__:  Explain the confidence interval in context of the research question:  
__ANSWER__:  


Create a 93% confidence interval for the __true population average__ number of siblings (`How_Many_Siblings`) of Brother Cannon's students($\mu_{\text{number of siblings}}$):  

```{r}


```

__QUESTION__:  Explain the confidence interval in context of the research question:  
__ANSWER__:  




# Medical Malpractice Lawsuits

Create a 99% Confidence interval for the __true population average__ payout for operating on the wrong patient (`Wrong_Patient`) in malpractice lawsuits ($\mu_{\text{payout}}$):

```{r}

malpractice <- import("https://github.com/byuistats/Math221D_Course/raw/main/Data/WrongSiteWrongPatient.xlsx")

```

__QUESTION__:  Explain the confidence interval in context of the research question:  
__ANSWER__:  





# Old Faithful


Create a 90% Confidence Interval for the __true population average__ wait time for an Old Faithful eruption ($\mu_{\text{wait time}}$):  

```{r}

old_faithful <- rio::import("https://byuistats.github.io/M221R/Data/old_faithful.xlsx")


```


__QUESTION__:  Explain the confidence interval in context of the research question:  
__ANSWER__:  


