---
title: One- and Two-sample Statistical Inference
subtitle: Application Activity
format:
  html:
    self-contained: true
    embed-resources: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```


# Introduction

This activity is designed to provide opportunities for practicing one- and two-sample hypothesis tests and confidence intervals.  

From context in the background information, you should be able determine what statistical test is appropriate and what the null and alternative hypotheses should be.  

For each problem:  

  - Carefully read the background information  
  - Define the null and alternative hypotheses  
  - Perform the relevant statistical test in R  
  - Check the test requirements  
  - Construct and interpret the confidence interval  

__NOTE__:  If not stated, use $\alpha=0.05$.  Also, if not specified, you can create graphs using base R or ggplot().  


Don't forget to load the libraries:

```{r}

library(tidyverse)
library(mosaic)
library(rio)
library(car)


```



# Problem 1 

A nutritionist claims that the average sodium content in a certain brand of canned soup is 900 mg per serving. A student takes a random sample of 29 cans and measures the sodium content.  The nutritionist wants to test if there is significantly different amount of sodium than advertised.  

Use the significance level, $\alpha = 0.10$

```{r}
##Sample Data (mg) 
sodium <- data.frame(mg = c(882, 915, 890, 908, 899, 883, 917, 905, 892, 879, 910, 886, 895, 901, 884, 897, 909, 894, 891, 900, 906, 882, 899, 913, 888, 892, 900, 904, 887))
```


State the hypotheses.

$$Ho:  $$  
$$Ha:  $$  

Perform the appropriate statistical test:  

```{r}

```


__QUESTION__:  What is the P-value?  
__ANSWER__:  

__QUESTION__:  State your conclusion in context of the research question.  
__ASNWER__:  


## Test Requirements

Check the test requirements for this hypothesis test:

```{r}

```


__QUESTION__:  What are the test requirements, and are they satisfied?  
__ANSWER__:  


## Confidence Interval

Calculate a 90% confidence interval for the population average sodium in the cans.  

```{r}

```

__QUESTION__:  Interpret the confidence interval in context of the research question.  
__ANSWER__:  


---

# Problem 2  

A researcher wants to compare the cholesterol levels (mg/dL) between patients taking a new drug and those taking a placebo. The researcher believes the new drug should lower cholesterol.  A group of 20 subjects are randomly assigned to a treatment or a placebo group.  

Run the following code chunk to create a dataset called cholesterol:  

__NOTE__:  There are some intermediate steps in the code below to create the final dataset called `cholesterol` that will be used for the analysis.  This dataset has 2 columns:  `treatment` and `Cholesterol`.  

```{r}
drug <- c(180, 185, 190, 195, 188, 182, 190, 193, 189, 185)
placebo <- c(188, 196, 196, 204, 197, 191, 201, 203, 202, 196)
treatment <- c(rep("Drug", length(drug)), rep("Placebo", length(placebo)))
cholesterol  <- data.frame(treatment = treatment,
                           Cholesterol = c(drug, placebo))

```


State the hypotheses.

$$Ho:  $$  
$$Ha:  $$  


## Summarize the Data

Create a table of summary statistics comparing cholesterol levels for each treatment group:

```{r}


```


__QUESTION__:  What is the difference between the sample averages of cholesterol between the 2 groups?  
__ANSWER__:  


Use `ggplot()` to create a side-by-side boxplot of the cholesterol levels of each treatment group: 

```{r}


```



## Hypothesis Test

Perform the appropriate statistical test:  

```{r}

```


__QUESTION__:  What is the P-value?  
__ANSWER__:  

__QUESTION__:  State your conclusion in context of the research question.  
__ASNWER__:  


### Test Requirements

Check the test requirements for this hypothesis test:

```{r}

```


__QUESTION__:  What are the test requirements, and are they satisfied?  
__ANSWER__:  


## Confidence Interval

Calculate a 95% confidence interval for the difference between the mean cholesterol levels:  

```{r}

```

__QUESTION__:  Interpret the confidence interval in context of the research question.  
__ANSWER__:  

---


# Problem 3

A dietitian measures the weight (kg) of 15 participants before and after a 4-week diet program.  

Run the following code chunk to create the dataset for this question:

```{r}
weight <- data.frame(subject = 1:15,
                     before = c(70, 65, 80, 75, 60, 68, 77, 69, 73, 66, 72, 71, 67, 74, 69),
                     after = c(70, 58, 79, 77, 57, 64, 78, 69, 69, 66, 69, 68, 63, 75, 68))

```


State the hypotheses.  

$$Ho:  $$  
$$Ha:  $$  


## Summarize the Data

Create histograms for before and after weights separately:  

```{r}

```

Create a histogram of the differences (after-before):

```{r}


```


__QUESTION__:  What is the average difference between after and before weights?  
__ANSWER__:  


## Hypothesis Test

Perform the appropriate statistical test:  

```{r}

```


__QUESTION__:  What is the P-value?  
__ANSWER__:  

__QUESTION__:  State your conclusion in context of the research question.  
__ASNWER__:  


### Test Requirements

Check the test requirements for this hypothesis test:

```{r}

```


__QUESTION__:  What are the test requirements, and are they satisfied?  
__ANSWER__:  


## Confidence Interval

Calculate a 95% confidence interval for the mean difference between before and after weights:  

```{r}

```

__QUESTION__:  Interpret the confidence interval in context of the research question.  
__ANSWER__:  


