---
title: "Confidence Intervals for a Mean"
---
```{r setup, include=FALSE}
# Ignore this bit of code.  This makes it so that you can render the qmd file even when there are errors in your code.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```

# Introduction

__Statistical Inference__ is the practice of using data sampled from a population to make conclusions about *population parameters*.  

The two primary methods of statistical inference are:  

  1. Hypothesis Testing
  2. Confidence Intervals
  
Hypothesis testing is the practice of gathering evidence in an attempt to disprove a hypothesis.  We either **reject** the conventional wisdom or **fail to reject**.  Proper scientific questions are falsifiable, meaning they can be disproven.  

Confidence intervals are a way to use data to estimate the unknown population parameter, often the population mean, $\mu$.  


This chapter lays the foundation for Hypothesis Testing.  


## Review:  distribution of sample means

Both methods of inference rely on the validity of the assumption that the distribution of the sample mean is normally distributed.  

Recall that the distribution of sample means is normal when:  

  1. The underlying population is normally distributed
  2. The sample size, n, is sufficiently large ($n<30$ for this class) for the Central Limit Theorem to apply




# Confidence Intervals

Confidence intervals are a way to estimate a population parameter without assuming a Null Hypothesis.  

## Confidence Interval

We can also calculate a confidence interval for the above example.  


Recall the formula for a confidence interval is, for a given $z^*$ associated with a desired level of confidence:

$$ CI = \bar{x} \pm z^*\frac{\sigma}{\sqrt{n}}$$

Recall the $z^*$ for common confidence levels:

```{r, fig.width = 9}
library(tidyverse)
library(pander)
tibble(`Conf. Level` = c(0.99, 0.95, 0.90), `Z*` = c(2.576, 1.96, 1.645)) %>% pander()
```


