---
title: "Quantitative Data Visualizations - Practice"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```

# Introduction

This is an opportunity for you to practice creating graphical summaries of quantitative data.  

We  will explore 2 datasets.  The first contains information about the duration and the time between geyser eruptions at Old Faithful in Yellowstone National park.  

The second is data collected about expressions of gratitude and their impact on subjective well-being.  

For each dataset, you will create numerical summaries of the data.  


# Load libraries and Data
```{r}

library(rio)
library(mosaic)
library(tidyverse)
library(car)


old_faithful <- rio::import("https://byuistats.github.io/M221R/Data/old_faithful.xlsx")

gratitude <- rio::import("https://byuistats.github.io/M221R/Data/gratitude.xlsx")

```

# Old Faithful

Create a histogram and a boxplot for Duration of Old Faithful eruptions:  

```{r}


```

__QUESTION__:  Describe the shape of the distribution of duration.  
__ANSWER__:  



# Gratitude

In this dataset, there is a column called `treatment` which has the group designation, and a single column called `happiness` which is the self-reported happiness score for the participants in the sample.  

## Histogram

Create a histogram and boxplot for happiness scores:  

```{r}


```

__Question__: What is the general shape of the distribution?  
__Answer__:  



