---
title: "Practice:  Bivariate Data"
subtitle: "Quantitative Response and Quantitative Explanatory Variables"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```


# Introduction

In this assignment, you will practice regression analysis including:

  * Plotting bivariate data
  * Calculating and interpreting the correlation coefficient, r
    
```{r}

library(tidyverse)
library(mosaic)
library(rio)
library(car)

```

# Car Prices and Mileage

You are interested in purchasing an all wheel drive Acura MDX for those slick Rexburg winters.  You found what you think is a good deal for on a low-mileage 2020 model but you'd like to be sure.  You go on Autotrader.com and randomly select 23 Acura MDX's and collect Price and Mileage information.

Load the data and use R to answer the questions below.  

```{r}

cars <- read_csv('https://github.com/byuistats/Math221D_Cannon/raw/master/Data/acuraMDX_price_vs_mileage.csv')

glimpse(cars)

```

__QUESTION__:  What is the response/dependent variable?  
__ANSWER__:  


__QUESTION__:  What is the explanatory variable?  
__ANSWER__:  


__QUESTION__:  What do you think is the nature of the relationship between the two?  (strength and direction)
__ANSWER__:  


__QUESTION__:  What is your best guess at the correlation coefficient, $r$?  
__ANSWER__:  


## Plot the Data and calculate r

```{r}

plot()

cor()

```


__QUESTION__:  Does the relationship look linear?  
__ANSWER__:  


__QUESTION__:  What does this r measure?  
__ANSWER__:  



# Manatee Deaths and Motorboat Registrations in Florida

Florida is a fabulous place for experiencing wildlife and recreation.  Unfortunately, sometimes those two activities conflict.

Researchers collected over 30 years of data about water craft registrations (motor and non-motor boats) and manatee deaths.  The goal of the research is to evaluate the relationship between boat registrations and manatee deaths.

## Load the data:

```{r}
manatees <- read_csv('https://github.com/byuistats/Math221D_Cannon/raw/master/Data/manatees.csv')

glimpse(manatees)
```


__QUESTION__:  What is the response/dependent variable?  
__ANSWER__:  


__QUESTION__:  What is the explanatory variable?  
__ANSWER__:  


__QUESTION__:  What do you think is the nature of the relationship between the two?  
__ANSWER__:  


__QUESTION__:  What is your best guess at the correlation coefficient, $r$?  
__ANSWER__:  


## Plot the Data and calculate r

```{r}



```

__QUESTION__:  Does the relationship look linear?  
__ANSWER__:  





# MCAT Score and GPA

The MCAT is an entrance exam for medical schools.  It seems likely that there is a relationship between your undergraduate GPA and how well you do on the MCAT.  

GPA and MCAT score data were collected on 55 prospective medical students.  

__QUESTION__:  What is the response/dependent variable?  
__ANSWER__:  


__QUESTION__:  What is the explanatory variable?  
__ANSWER__:  


__QUESTION__:  What do you think is the nature of the relationship between the two?  
__ANSWER__:  


__QUESTION__:  What is your best guess at the correlation coefficient, $r$?  
__ANSWER__:  


## Plot the Data and calculate r

```{r}



```

__QUESTION__:  Does the relationship look linear?  
__ANSWER__:  


