---
title: "Practice: Summarizing Categorical Data"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```

# Introduction

Reasons why patients visit a chiropractor vary and may depend on where you live.  

Use the chiropractic dataset to practice summarizing categorical data.  

## Load the Libraries and Data

```{r}

library(tidyverse)
library(mosaic)
library(rio)
library(car)

chiropractic <- import('https://raw.githubusercontent.com/byuistats/Math221D_Cannon/master/Data/chiropractic_data.csv')

```


# Why Visit?  

__QUESTION__:  List all the levels for `motivation` in the data:  
```{r}
unique()
```


Create a sorted bar chart for reasons why people see a chiropractor:  

```{r}


```

__QUESTION__:  What is the most common motivation for visiting a chiropractor?  
__ANSWER__:  


# Regional Differences? 

Create and name a contingency table that counts the motivation by region.  
```{r}
chiro_tbl <- table()

prop.table()

```

__HINT__:  Proportions should sum to 1 for each region. Which margin you use will depend on how you defined your table above.  

__QUESTION__:  What percent of Australians visit a chiropractor for Wellness?  
__ANSWER__:  

__QUESTION__:  What percent of Europeans visit a chiropractor for Wellness?  
__ANSWER__:  

# Visualize

Use the `chiro_tbl` created above to create a grouped bar chart:  

```{r}

```


Swap the row and column table and recreate the graph above with the opposite grouping:  

```{r}

chiro_tbl2 <- table()

barplot()

```

__QUESTION__:  In your opinion, which visualization tells a more compelling story?  Why?  
__ANSWER__:  

