---
title: "Confidence Intervals and Hypothesis Tests For Proportions - Practice"
format:
  html:
    self-contained: true
    embed-resources: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```



# Introduction

In this exercise, you will practice creating and interpreting confidence intervals for proportions.  


# Bird Watching

A wildlife researcher is studying a rare bird species in a southeast Idaho.  They identify a sample of 500 birds, and only 6 are found to be of this rare species. Estimate the proportion of the entire bird population that is this rare species using a 95% confidence interval.  



```{r}

prop.test(x = , n = , conf.level=)$conf.int

```


__QUESTION__:  Interpret the confidence interval in context of the research question:  
__ANSWER__:  


__QUESTION__:  Are the assumptions for the normality of $\hat p$ are met?  (Check $n\hat{p}$ and $n(1-\hat{p})$)  
__ANSWER__:  

__QUESTION__:  Perform a statistical hypothesis test to see if the population proportion, $p$, is less than 3%.  Use $\alpha=0.05$.  

```{r}

prop.test(x=, n=, p=, alternative=)

```


__QUESTION__:  What is the p-value for the above test?  
__ANSWER__:  


__QUESTION__:  State your conclusion using the technical requirements and in context of the research question.  
__ANSWER__:  

__QUESTION__:  Are the test requirements for a statistical hypothesis for a proportion met?  (Check $np$ and $n(1-p)$)  
__ANSWER__:  



# Customer Satisfaction Survey

A company surveyed 150 customers about their satisfaction, and 105 responded that they were satisfied. Estimate the proportion of all customers who are satisfied with the company's service using a 99% confidence level.  



```{r}

```


__QUESTION__:  Interpret the confidence interval in context of the research question:  
__ANSWER__:  


__QUESTION__:  Are the assumptions for the normality of $\hat p$ are met?  Why?  
__ANSWER__:  


# Health Study

A health study surveyed 250 individuals about their exercise habits, and 140 reported exercising regularly. Construct a 97.25% confidence interval for the proportion of the general population that exercises regularly.



```{r}

```


__QUESTION__:  Interpret the confidence interval in context of the research question:  
__ANSWER__:  


__QUESTION__:  Are the assumptions for the normality of $\hat p$ are met?  Why?  
__ANSWER__:  



