---
title: "Data Cleaning Practice"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```


# Notes:  

Break this into 3, then do cumulative retrieval (pair-share each day; start with the hardest one so it gets the most retrieval); possibly have some scaffolding slide, prompt words.  

Groups of 3 for a week; what if you used the same topic?  Do 3 filter exercises; do 3 gg-plots, etc.  

# Introduction

In this exercise, you'll practice the basics of data cleaning and visualization.  

Load and explore the data then respond to the questions below.  


```{r}
library(tidyverse)
library(mosaic)
library(rio)
library(car)

real_estate <- read_csv('https://github.com/byuistats/Math221D_Course/raw/refs/heads/main/Data/MadisonCountyRealEstateA.csv')

glimpse(real_estate)

```

__QUESTION__:  Make a Histogram of `ListPrice`

```{r}

ggplot(real_estate, aes(x = ListPrice)) +
  geom_histogram()

```

__QUESTION__:  What is the shape of the distribution of list prices?  
__ANSWER__:  


__QUESTION__:  Make a table of summary statistics for `ListPrice`

```{r}

favstats(real_estate$ListPrice)

```



__QUESTION__:  What is the minimum value for list prices?  
__ANSWER__:  1.47



Create a clean dataset that:

  1. Removes houses that are more than $1M
  2. Has only 1 Story houses
  3. No Missing Elementary school information
  4. Is in Rexburg
  5. Contains only the columns: ListPrice, Style, Elementary, City, and IsFixerUpper
  
  
```{r}

unique(real_estate$City)

clean <- real_estate %>%
  filter(ListPrice > 2,
         ListPrice<1000000,
         Style == "1 Story",
         Elementary != "MISSING",
         City == "REXBURG") %>%
  select(ListPrice, Style, Elementary, City, IsFixerUpper)

clean

```


__QUESTION__:  What percent of homes in the clean dataset are fixer-uppers?

```{r}

prop.table(table(clean$IsFixerUpper))

```

__QUESTION__:  Use GGPLOT to create a boxplot for `ListPrice` that compares Elementary Schools?  


```{r}

ggplot(clean, aes(y = ListPrice, x = fct_reorder(Elementary, ListPrice), fill=Elementary)) +
  geom_boxplot() +
  theme_bw() +
  theme(
    legend.position = "null"
  )

```


__QUESTION__:  Which Elementary school appears to have the lowest distribution of housing prices?  
__ANSWER__:  
