---
title: "Data Cleaning Practice"
format: html
---

# Introduction

Take 10 minutes to:

  1. Load the data
  2. Review the data
  3. Determine the response variable
  4. Pick an explanatory variable
  5. Make a clean dataset that has only your response and explanatory variable and has removed bad data (using `filter()` and `select()`)
  6. Make a graph that visualizes the relationship between the 2



```{r}
library(tidyverse)
library(mosaic)
library(rio)
library(car)

real_estate <- read_csv('https://github.com/byuistats/Math221D_Course/raw/refs/heads/main/Data/MadisonCountyRealEstateA.csv')

```

__QUESTION__:  Make a Histogram of `ListPrice`

```{r}


```


__QUESTION__:  Make a table of summary statistics for `ListPrice`

```{r}


```



Create a clean dataset that:

  1. Removes houses that are more than $1M
  2. Has only 1 Story houses
  3. No Missing Elementary school information
  4. Is in Rexburg
  5. Contains only the columns: ListPrice, Style, Elementary, City, and IsFixerUpper
  
  
```{r}


```


__QUESTION__:  What percent of homes in the clean dataset are fixer-uppers?

```{r}



```

__QUESTION__:  Use GGPLOT to create a boxplot for ListPrice that compares Elementary Schools?  




