---
title: "Data Cleaning Practice"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```

# Introduction

In this exercise, you'll practice the basics of data cleaning and visualization.  

Load and explore the data then respond to the questions below.  


```{r}
library(tidyverse)
library(mosaic)
library(rio)
library(car)

real_estate <- read_csv('https://github.com/byuistats/Math221D_Course/raw/refs/heads/main/Data/MadisonCountyRealEstateA.csv')

```

__QUESTION__:  Make a Histogram of `ListPrice`

```{r}


```

__QUESTION__:  What is the shape of the distribution of list prices?  
__ANSWER__:  


__QUESTION__:  Make a table of summary statistics for `ListPrice`

```{r}


```



__QUESTION__:  What is the minimum value for list prices?  
__ANSWER__:  



Create a clean dataset that:

  1. Removes houses that are more than $1M
  2. Has only 1 Story houses
  3. No Missing Elementary school information
  4. Is in Rexburg
  5. Contains only the columns: ListPrice, Style, Elementary, City, and IsFixerUpper
  
  
```{r}


```


__QUESTION__:  What percent of homes in the clean dataset are fixer-uppers?

```{r}



```

__QUESTION__:  Use GGPLOT to create a boxplot for `ListPrice` that compares Elementary Schools?  


```{r}



```


__QUESTION__:  Which Elementary school appears to have the lowest distribution of housing prices?  
__ANSWER__:  
