---
title: One and Two Sample Proportion Practice
format:
  html:
    self-contained: true
    embed-resources: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)
```


# One-Sample Test of Proportions

## Teenage Smoking Habits

A study was conducted to determine the proportion of American teenagers between 13 and 17 who smoke. A survey from 10 years ago claimed that 15% percent of all teenagers smoke. 

A new Gallup survey interviewed a nationally representative sample of 785 teenagers aged 13 to 17. Seventy-one (71) teenagers in the survey acknowledged having smoked at least once in the past week. 

We want to see if the new study shows a *decrease* in the  percentage of teenagers who smoke from the 15% reported a decade ago.  

Perform the appropriate hypothesis test and create a confidence interval for the true proportion of teenagers who smoke.

### Hypothesis Test

State the null and alternative hypotheses:  

$$H_0: $$  
$$H_a: $$
Choose your confidence level:

$$\alpha = 0.$$  

Perform the appropriate analysis:  

```{r}

prop.test(x = , n = , alternative = "")

```

__Question__:  What is the P-value?  
__Answer__:  

__Question__:  Based on your chosen $\alpha$ and P-value, what is your conclusion?  
__Answer__:  


#### Requirements

Recall that we must check that we have a big enough sample size to trust our p-value.  To do this, we check that there are at least 10 expected "success" and "failures" for a given sample size, n:  

Check:  

$$np \ge 10$$

$$n(1-p) \ge 10$$
```{r}
# Fill in n and the hypothesized p
n <- 
p <- 

n*p >= 10

n*(1-p) >= 10


```


__QUESTION__:  Are the requirements for a hypothesis test satisfied?  
__ANSWER__:  


### Confidence Interval

Create a 99\% confidence interval for the true population proportion of teenagers who smoke?  

```{r}

```


__QUESTION__:  Interpret the confidence interval in context of the research question:  
__ANSWER__:  


#### Check Requirements

Recall that Confidence Intervals do not depend on a hypothesized proportion, so the requirements are a little different.  For Confidence Intervals we check:  

$$n\hat{p} \ge 10$$

$$n(1-\hat{p}) \ge 10$$

```{r}
# Fill in X and N and check that there are enough "successes" and "failures"

x <- 
n <- 
phat <- x/n

n*phat >= 10

n*(1-phat) >= 10
```



# 2-Sample Proportion Tests

## A Penny Saved?

A random sample of BYU-Idaho students was surveyed and asked if they were in favor of retaining the penny as a form of currency in the United States. Out of the 116 women surveyed, 80 said that they were in favor of retaining the penny as a form of currency. Of the 137 men surveyed, 91 said that they were in favor of retaining the penny. 

For these questions, let group 1 represent women and group 2 represent men. 


### Hypothesis Test

Test to see if there is a *difference* between the proportion of women who want to keep the penny and the proportion of men who want to keep the penny. Use a level of significance of $\alpha = 0.05$.

State your null and alternative hypotheses (replace the question marks with the appropriate symbols):  

$$H_0: p_{female} ???  p_{male}$$  
$$H_a: p_{female} ??? p_{male}$$  

Perform the appropriate test:  
```{r}

prop.test(x = c(), n = c(), alternative = "")

```


__Question__:  What is the P-value?  
__Answer__:  


__Question__:  Based on $\alpha = 0.05$, state your conclusion in context of the research question:  
__Answer__:  



### Confidence Interval

Create a 95\% confidence interval for the *difference* in the proportion of females to males who prefer to keep the penny:

```{r}

```

__Question__:  Interpret the confidence interval in context of the research question:  
__Answer__:  


### Test Requirements

Recall that the requirements for Hypothesis Testing and Confidence Intervals for 2-sample proportions are the same.  We must check that there are more than 10 "successes" and "failures" in *both* samples:  

$$ n_1\hat{p}_1 \ge 10$$
$$n_1(1-\hat{p}_1) \ge 10$$
$$ n_2\hat{p}_2 \ge 10$$
$$n_2(1-\hat{p}_2) \ge 10$$


Use the following calculator to check the above requirements:  

```{r}

# All must be true:

x1 <- 
n1 <- 
phat1 <- x1/n1

n1*phat1 >= 10
n1*(1-phat1) >=10

x2 <- 
n2 <- 
phat2 <- x2 / n2

n2*phat2 >= 10
n2*(1-phat2) >=10


```


__Question__:  Are the requirements satisfied for assuming normality?  
__Answer__:  



# Couples Retreat?

A marriage counselor conducted a study of couples, categorizing each of the couples as "communicative" or "non-communicative". Among other things, the counselor wanted to see whether the percentage of communicative couples whose marriage ended in separation or divorce was *less than* the percentage of non-communicative couples whose marriage ended in separation or divorce. 

Of the 97 couples classified as "communicative", 17 ended in separation or divorce.  Of the 56 couples classified as "non-communicative", 13 ended in separation or divorce.  

## Hypothesis Test

Construct a null and alternative hypothesis for the study:

$$H_0: $$  
$$H_a: $$  

Perform the appropriate analysis:  

```{r}
prop.test()
```


__Question__:  What is the P-value?  
__Answer__:  


__Question__:  Based on your decision rule, state your conclusion in context of the research question:   
__Answer__:  


## Confidence Interval

Create a 95\% confidence interval for the difference in the proportion of divorces between communicative and non-communicative couples:

```{r}

```

__Question__:  Interpret the confidence interval in context of the question:  
__Answer__:  


### Test Requirements

Use the following calculator to check the requirements for the hypothesis test and confidence interval:  

```{r}

# All must be true:

x1 <- 
n1 <- 
phat1 <- x1/n1

n1*phat1 >= 10
n1*(1-phat1) >=10

x2 <- 
n2 <- 
phat2 <- x2 / n2

n2*phat2 >= 10
n2*(1-phat2) >=10


```



__QUESTION__:  Are the requirements for the hypothesis test and confidence interval satisfied?  
__ANSWER__:  

